Optimizing Abstract Abstract Machines

### 2012-11-12 paper/paper.tex
Balance.
Touching up prose.

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
For the duration of this paper, $\sqcup$ is a point-wise lifting of $\cup$:
\begin{align*}
\msto\sqcup \varnothing &= \msto \\
\msto\sqcup \msto'[\maddr \mapsto S] &= \lambda \maddralt. \left\{
  \begin{array}{ll}
    \msto(\maddr)\cup S & \mbox{if } \maddralt = \maddr, \maddr\in dom(\msto) \\
    S & \mbox{if } \maddralt = \maddr, \maddr\notin dom(\msto) \\
    (\msto\sqcup\msto')(\maddralt) & \mbox{otherwise}
  \end{array}\right.
\end{align*}

@@

\newpage
\section{Baseline analyzer}
\label{sec:baseline}

@@

The final approximation we make to get to our baseline analysis is
store-allocating results of application sub-expressions. Notice that
the $\kfn{}$ continuation stores a value - this makes the space of
continuations quadratic rather than linear in 0-CFA. We instead
allocate an address for the value position when we create the
continuation. This address and the tail address are both determined by
the label of the application point, so the space becomes linear and
the overall complexity drops to cubic. Note that this is a critical
abstraction in languages with n-ary functions, since otherwise the
continuation space becomes ${\mathcal O}(n^n)$!

@@

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
either ANF or CPS. We can also play a representation trick and remove
$\menv$ since we know it is the identity function in 0-CFA. We will
refer to this change in the semantics as value indirection.

@@

We discuss a variety of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation. They
come in two flavors: decreasing the size of the represented state
space [SS], and decreasing intermediate data structures [DS].

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector in
   the 0-CFA case. (preallocation)}
 \item{[SS] coarsely abstract literal data (abstracting compound data)}
\end{itemize}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation/fixpoint, and each improves performance by
integer factors. The first three remain in the purely functional
realm. The final three use mutation in small, localized places, so
they do not increase conceptual complexity drastically.

@@

A shortcoming of the AAM approach is that much of the non-determinism
is spurious. For example, in a function application, {\tt (f x y)},
where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
only to be joined back together in their respective application positions. We would
see transitions of the following shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

@@

This semantics introduces a subtle precision difference over the
baseline that is negligible. Consider a configuration where a
reference to a variable and a binding of a variable will happen in one
step. With laziness, the reference will mean the original binding(s)
of the variable or the new one, because the actual store lookup is
delayed precisely one step (i.e. laziness is administrative). Without
laziness, the reference will fan out to all the bindings of the
variable before the new binding happens and thus might have an
observable precision difference. In our experience, this situation
happens too rarely to warrant the significant increase in time and
memory needed for this eager non-determinism. Indeed, were the
variable reference a step later and another binding is not made in
that step, the results of the two approaches are the same.

The administrative nature of laziness means that we could remove the
loss in precision by essentially duplicating the reduction relation to
specialize variable lookup. This works since in the semantics of ISWIM
with value indirection consumes the value component of states in
one step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

@@

We can eliminate the {\tt ev} states from the run-time interpretation
of a program by specializing the machine transition relation to the
program being analyzed. This eliminates interpretative overhead by
first compiling the program into ``byte code'' instructions.

@@

Each step in the fixpointer for the above techniques requires joining
several, commonly large, stores together. Even with immutable hashes
that utilize sharing, this is memory-intensive. Worse, this
contributes an obscene fraction of the runtime. If we instead track
the changes to the store made over all steps and afterward update the
store, we end up with much less allocation and computational overhead.

@@

\begin{align*}
bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
               \maddr &= \mvar\mcntr' \\
(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
\end{align*}

We also lift $\machstep$ differently to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated.

@@

To ensure termination, we guard adding $c$ to {\tt todo} by the following check:
\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

After all steps complete, we apply $\msdiff$ to $\msto$ and increase
{\tt T} as long as there was at least one change in $\msdiff$. This
logic is only sound if we know that each $\ttuple{\maddr}{\mval{s}}$
in $\msdiff$ would change the value of $\maddr$ in the current
store. Thus, we also guard additions to $\msdiff$ so that only updates
that would change the store are permitted. Each time $\msdiff$ is
successfully extended, we set {\tt $\msdiff$?} to true. Before each
individual step, we set it to false.

@@

This technique by itself only works in 0-CFA, since there are many more
addresses that can be allocated in 1-CFA and higher. To support
general $k$-CFA in a vector representation, we use a growable vector
and a map from addresses to assigned indices. When assignable indices
run out, we grow the vector. The cost of this map is less than a hash
representation for the whole store since typically there are more
references to an address than allocations of it. This is more
space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
up-front and addressing by the contour converted to polynomial, since
many combinations of labels are not likely to occur. The all up-front
allocation again breaks down when we add compound data, which
introduce new kinds of labels (e.g. left and right components of a
tuple).

@@

Practical issues of performance arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data. Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
value of which do not particularly matter for control-flow
analysis. If we interpret a list literal of 30 elements (that again
have 10 elements each) as 300 applications of {\tt cons}, and then
interpret a function that traverses the tree, we will end up
interpreting that function at least 300 times, when once or twice
would do.

We will first explain the natural abstraction of tuples, and then we
will explain a less precise allocation strategy from uniform $k-CFA$
that we use for large compound data literals that we modified from the
implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.

@@

Upon evaluating a {\tt cons} application, we instead allocate two
addresses $a$ and $d$, join them to the respective values in the
store, and return the flattened {\tt (cons $a$ $d$)} value. Since
these addresses are all distinguished at different syntactic callsites
in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
for a sequence of calls to {\tt cons}, this abstraction explodes the
value space. Analyzing a function that counts the number of atoms in a
literal s-expression would actually interpret that function at least
that number of times (more because of intermediate conses). Indeed,
even in our fastest parameterization, this abstraction causes the
analysis of Boyer to be 430 times slower than the approach we will now
describe.

>>>>

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
For the duration of this paper, since our example abstract value space
has no interesting ordering, $\sqcup$ is a point-wise lifting of
$\cup$:
\begin{align*}
\msto\sqcup \varnothing &= \msto \\
\msto\sqcup \msto'[\maddr \mapsto S] &= \lambda \maddralt. \left\{
  \begin{array}{ll}
    \msto(\maddr)\cup S & \mbox{if } \maddralt = \maddr, \maddr\in dom(\msto) \\
    S & \mbox{if } \maddralt = \maddr, \maddr\notin dom(\msto) \\
    (\msto\sqcup\msto')(\maddralt) & \mbox{otherwise}
  \end{array}\right.
\end{align*}

@@

\newpage
\section{Reduction semantics to baseline analyzer}
\label{sec:baseline}

The uniform $k$-CFA allocation strategy would make the fixpoint of the
reduction relation in figure \ref{fig:aam} a computable analysis, but
not an effective one. This is not the strategy that AAM, nor we,
recommend. We explain a succession of approximations to reach the
baseline analysis that we compare performance against. But first, we
need a fixpoint combinator to drive the succession of reduction
relations we define.

@@

The final approximation we make to get to our baseline analysis is
store-allocating results of application sub-expressions. The semantics
of the previous section is where AAM leaves off. However, notice that
the $\kfn{}$ continuation stores a value - this makes the space of
continuations quadratic rather than linear in 0-CFA. A linear space
drops the complexity to cubic.

To achieve this, we instead allocate an address for the value position
when we create the continuation.  We will refer to this change in the
semantics as store-allocated results. This address and the tail address are
both determined by the label of the application point, so the space
becomes linear and the overall complexity drops to cubic. Note that
this is a critical abstraction in languages with n-ary functions,
since otherwise the continuation space becomes ${\mathcal O}(n^n)$!

@@

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
ANF. We can also play a representation trick and remove
$\menv$ since we know it is the identity function in 0-CFA.

@@

We discuss a succession of techniques for increasing performance by
decreasing the interpretive overhead of abstract interpretation. They
come in two flavors: decreasing the size of the represented state
space [SS], and decreasing the amount of intermediate data structures [DS].

\begin{itemize}
 \item{[SS] quotient by sets of values (lazy non-determinism)}
 \item{[SS] remove {\tt ev} states (abstract compilation)}
 \item{[DS] operate on store changes rather than whole stores (store deltas)}
 \item{[DS] represent the store as a global, mutable hash and
   determine a state's store by the number of changes made to the
   global store at the time of visiting (timestamp)}
 \item{[DS] use a global mutable workset}
 \item{[DS] represent the global store as a global, mutable vector (preallocation)}
 \item{[SS] coarsely abstract literal data (abstracting compound data)}
\end{itemize}

Some techniques preserve the precision of the underlying analysis, and
others do not. We will discuss the design decisions behind the
precision-lossy techniques.  Each technique requires minimal changes
to the reduction relation/fixpoint, and each improves performance by
integer factors. The first three and last remain in the purely
functional realm. The final three use mutation in small, localized
places, so they do not increase conceptual complexity
drastically. Also, preallocation and mutable workset are standard
implementation practices \cite{ianjohnson:Shivers:1991:CFA}.

@@

A shortcoming of the AAM approach is that much of the non-determinism
is spurious. For example, in a function application, {\tt (f x y)},
where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt f}, {\tt x} and {\tt y} (or in the our baseline analysis, the sum),
only to be joined back together in their respective application positions. We would
see transitions of the following shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

@@

\paragraph{Precision loss} This semantics introduces a subtle precision difference over the
baseline that is negligible. Consider a configuration where a
reference to a variable and a binding of a variable will happen in one
step. With laziness, the reference will mean the original binding(s)
of the variable or the new one, because the actual store lookup is
delayed precisely one step (i.e. laziness is administrative). Without
laziness, the reference will fan out to all the bindings of the
variable before the new binding happens and thus might have an
observable precision difference. In our experience, this situation
happens too rarely to warrant the significant increase in time and
memory needed for this eager non-determinism. Indeed, were the
variable reference a step later and another binding not made in
that step, the results of the two approaches are the same.

\paragraph{Avoiding precision loss} The administrative nature of laziness means that we could remove the
loss in precision by essentially duplicating the reduction relation to
specialize variable lookup. This works since in the semantics of ISWIM
with store-allocated results consumes the value component of states in
one step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

@@

Laziness removes several represented states that would then have to be
stepped by the fixpoint combinator.  We can further eliminate the {\tt
  ev} states from the run-time interpretation of a program by
specializing the machine transition relation to the program being
analyzed. This eliminates interpretative overhead by first compiling
the program into ``byte code'' instructions.

@@

Each step in the fixpoint for the above techniques requires joining
several, commonly large, stores together. Even with immutable hashes
that utilize sharing, this is memory- and time-intensive. Worse, this
contributes an obscene fraction of the runtime. If we instead track
the changes to the store made over all steps and afterward update the
store, we end up with much less allocation and computational overhead.

@@

\begin{align*}
(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
\end{align*}
\begin{align*}
bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
               \maddr &= \mvar\mcntr'
\end{align*}

We also lift $\machstep$ to accommodate for this asymmetry
in the input and output. For each state that is stepped, we feed the
output changes to the next so that all changes get accumulated.

@@

The new fixpoint calculation is defined in
figure~\ref{fig:imperative}.  To ensure termination, we guard adding
$c$ to {\tt todo} by the following check:
\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

After all steps complete, we apply $\msdiff$ to $\msto$ imperatively and increase
{\tt T} as long as there was at least one change in $\msdiff$. This
logic leads to termination if we know that each $\ttuple{\maddr}{\mval{s}}$
in $\msdiff$ would change the value of $\maddr$ in the current
store. Thus, we also guard additions to $\msdiff$ so that only updates
that would change the store are permitted. Each time $\msdiff$ is
successfully extended, we set {\tt $\msdiff$?} to true. Before each
individual step, we set it to false.

@@

This technique by itself only works in 0-CFA, since there are many more
addresses that can be allocated in 1-CFA and higher. To support
general $k$-CFA in a vector representation, we use a growable vector
and a map from addresses to assigned indices. When assignable indices
run out, we grow the vector. The cost of this map is less than a hash
representation for the whole store since typically there are more
references to an address than allocations of it. This is more
space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
up-front and addressing by the contour converted to polynomial, since
many combinations of labels are not likely to occur. The all up-front
allocation again breaks down when we add compound data, which
introduce new kinds of labels (e.g. left and right components of a
tuple at the contour of creation).

@@

Practical issues of performance arise when one chooses the
``obvious,'' or ``natural'' abstraction for compound data. Our running
example will be the classic Boyer benchmark. The detail of note for
the Boyer benchmark is that it has large list literals, the exact
values of which do not particularly matter for many analyses. If we
interpret a list literal of 30 elements (that again have 10 elements
each) as 300 applications of {\tt cons}, and then interpret a function
that traverses the tree, we will end up interpreting that function at
least 300 times, when once or twice would do.

We will first explain the natural abstraction of tuples, and then we
will explain a less precise allocation strategy from uniform $k$-CFA
that we use for large compound data literals that we modified from the
implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.

@@

Upon evaluating a {\tt cons} application, we instead allocate two
addresses $a$ and $d$, join them to the respective values in the
store, and return the flattened {\tt (cons $a$ $d$)} value. Since
these addresses are all distinguished at different syntactic callsites
in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
for a sequence of calls to {\tt cons}, this abstraction explodes the
value space. Analyzing a function that counts the number of atoms in a
literal s-expression would actually interpret that function at least
that number of times (more because of intermediate conses). Indeed,
even in our fastest implementation, this abstraction causes the
analysis of Boyer to be 430 times slower than the approach we will now
describe.
